Versioned data: why it is needed and how it can be achieved (easily and cheaply)

### 2017-10-17 ms.tex
some edits throughout the text (also, i think we should avoid using an acronym for versioned data)

Many, perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time. Typical changes include: the adding new data, improving the quality of existing data, integrating with other databases, or re-structuring the database content to be able to address new questions. For example, a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while past versions of the dataset would remain accessible to ensure reproducibility.

@@

In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing Versioned Data (\textsc{vd}). Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer. With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer. In either the development of software or data, successive versions are released over time (called ``releases''). The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data. Importantly, our \textsc{vdd} systems uses well-established tools (Table \ref{tab:technologies}), ensuring high-level performance and stability.

@@

To aid reproducibility and efficient usage, many users will want access to all versions of any particular dataset programmatically (Table \ref{tab:user_requirements}). Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle. To make it easier for users to access \textsc{vd} via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \citep{R-2017}.

Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that facilitates access to releases of any \textsc{vd} hosted on \texttt{GitHub} (Fig. \ref{fig:technology_stack}). Specifically, the  \texttt{datastorr} package:
\begin{enumerate}
  \item Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub}.
  \item Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
\end{enumerate}
Using \texttt{datastorr}, a researcher can create and distribute a custom \textsc{R} package that facilitates access to their data with (very) minimal computational skills.

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}. The R package \texttt{baad.data} consists of only a few simple functions, that were automatically generated (along with the associated help files) with \texttt{datastorr}. For a user, accessing a version of the data is a simple as typing a single line of code (Fig. \ref{fig:technology_stack}). Accessing a different version of the data involves changing only the version number. From the users perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \citep[e.g.][]{Duursma-2016,Falster-2016}.

Using \texttt{datastorr}, dataset \emph{maintainers} can set up their own R package to deliver there versioned dataset simply by providing:
\begin{enumerate}
  \item a \texttt{GitHub} repository name (e.g., ``traitecoevo/baad.data'') where releases are stored
  \item the filename in the release that will contain the \textsc{vd}
  \item the function used to load this into \texttt{R}.
\end{enumerate}
A full tutorial explaining precisely how to set this up is available at \smurl{github.com/ropenscilabs/datastorr}.

@@

%work on this paragraph more
We strongly suspect that the issues we have identified in this article will not come as news to most readers: datasets are constantly evolving (hopefully improving) and despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact. In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought about versioning and developed specific solutions. But so far these as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences). But for an individual research group working with a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter. As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.

% Moved from above, need to integrate better
We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users. And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
% Reword next senetnce
However, we can imagine alternative methods for interacting with \texttt{GitHub} and building templates and adopting a \textsc{vdd} system.  The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.

Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many current platforms for storing and distributing versioned controlled data for low or no cost.  The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}. Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.  The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.  We argue that by linking together the tools described above, researchers can now produce, curate, and distribute \textsc{vd} much more easily and cheaply than in the past.

@@

Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently\citep{Fli, Dat}.
% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future

>>>>

Many, and perhaps most, of the datasets published in \emph{Scientific Data} are likely to be living entities, meaning the state-of-the-art (or ``canonical'') version of the dataset will change through time. Typical changes include: the adding new data, improving the quality of existing data, integrating with other databases, or re-structuring the database content to be able to address new questions. For example, a dataset on a biological organisms might be expanded through the addition of new records or improved through the correction of spelling mistakes in taxonomic names. As research around a data product grows, there might be many such additions. Ideally, any changes to a dataset would become immediately available to any interested parties, while past versions of the dataset would remain accessible to ensure reproducibility.

@@

In brief, the \textsc{vdd} workflow we present borrows best-practices for software development \citep{Perez-Riverol-2016} and applies them to the challenge of maintaining and distributing Versioned Data. Software developers maintain a core set of code which produces a binary executable file that can be installed on a users local computer. With a \textsc{vdd} system, data developers similarly maintain a core set of files (the ``code''), which produce an organised dataset that can be ``installed'' on a user's local computer. In either the development of software or data, successive versions are released over time (called ``releases''). The similarity in workflow between software and data then allows us to deploy the same technological platforms that are used in software development, for the development and distribution of data. Importantly, our \textsc{vdd} systems uses well-established tools (Table \ref{tab:technologies}), ensuring high-level performance and stability.

@@

To aid reproducibility and efficient usage, many users will want access to all versions of any particular dataset programmatically (Table \ref{tab:user_requirements}). Code to access a stream of \texttt{GitHub} releases could be written individually by each user, but this creates an unnecessary technological hurdle. To make it easier for users to access Versioned Data via code, we offer a novel implementation of \textsc{vdd} focussing on the \texttt{R} platform, as one of the most prominent platforms for data science \citep{R-2017}.

Here we introduce a dedicated \textsc{R} package, called \texttt{datastorr} (\smurl{github.com/richfitz/datastorr}), that facilitates access to releases of any Versioned Data hosted on \texttt{GitHub} (Fig. \ref{fig:technology_stack}). Specifically, the  \texttt{datastorr} package:
\begin{enumerate}
  \item Constructs the shell of a second, dataset-specific \textsc{R} package, which is used to access releases from a specific repository stored on \texttt{GitHub};
  \item Contains the main code needed to interact with the \texttt{GitHub} \textsc{api} to retrieve versions of the dataset.
\end{enumerate}
Using \texttt{datastorr}, a researcher can create and distribute a custom \textsc{R} package that facilitates access to their data with (very) minimal computational skills.

For example, \texttt{datastorr} was used to build the package \texttt{baad.data} (\smurl{github.com/traitecoevo/baad.data}), which is an interface to the Biomass and Allometry Database (\textsc{BAAD}) \citep{Falster-2015} stored at \smurl{github.com/dfalster/baad}. The R package \texttt{baad.data} consists of only a few simple functions and associated help files, that were automatically generated with \texttt{datastorr}. For a user, accessing a version of the data is a simple as typing a single line of code (Fig. \ref{fig:technology_stack}). Accessing a different version of the data involves changing only the version number. From the users perspective, the existence of the \texttt{baad.data} and \texttt{datastorr} packages makes reproducing analyses using specific versions of the data possible \citep[e.g.][]{Duursma-2016,Falster-2016}.

Using \texttt{datastorr}, dataset \emph{maintainers} can set up their own R package to deliver there versioned dataset simply by providing:
\begin{enumerate}
  \item a \texttt{GitHub} repository name (e.g., ``traitecoevo/baad.data'') where releases are stored;
  \item the filename in the release that will contain the Versioned Data;
  \item the function used to load this into \texttt{R}.
\end{enumerate}
A full tutorial explaining precisely how to set this up is available at \smurl{github.com/ropenscilabs/datastorr}.

@@

%work on this paragraph more
We acknowledge that the issues we have identified in this article will not come as news to most readers: datasets are constantly improving and, despite tremendous advances in data sharing and associated technologies over the last decade, there is little consensus about what to do with this fact. In some individual cases, such as \texttt{GenBank} (see \url{https://www.ncbi.nlm.nih.gov/genbank/sequenceids/}), researchers have thought about versioning and developed specific solutions. But so far these as we mentioned above, these datasets tend to be those managed by large, well-funded institutions or consortia and involve standard data types (e.g., sequences). But for an individual research group working with a specialized type of data, creating their own \texttt{GenBank} is clearly a non-starter. As a result, we argue that there is a substantial need for a easy, cheap, and flexible solution and suggest that adopting a \textsc{vdd} system may be one way forward.

% Moved from above, need to integrate better
We have created \texttt{datastorr} to make \textsc{vdd} easy to implement and manage for \texttt{R} users. And as we stated above, this has already been used to release a variety of datasets (Table \ref{tab:examples}).
% Reword next senetnce
But the design choices we made in \texttt{datastorr} represent only one of many possible ways to adopt a \textsc{vdd} system; we can imagine many alternative models for interacting with \texttt{GitHub} (or an entirely different service and technology) and accessing versioned data programmatically.  The key is not the specific technology, but rather the concept of creating, maintaining, and distributing versioned data.

Many of the key roadblocks preventing a switch from a static to a dynamic data world were technological: in the past, it took great deal of money and expertise to set up and maintain a \textsc{vdd} system, but now there are many platforms for storing and distributing versioned controlled data with little or no cost.  The website \texttt{GitHub} (\smurl{www.github.com}) is a commercial platform for hosting and interacting with \texttt{git} repositories. In practice, each dataset should be a separate \texttt{git} repository on \texttt{GitHub} (see Table \ref{tab:examples} for examples). Although mainly used for computer code, \texttt{GitHub} is now also being use to manage scientific data \citep{Perkel-2016}. Maintaining the version controlled data in the cloud has two main benefits: first, it provides a platform for multiple data contributors to sync their files and correspond about changes in the dataset, and second, it hosts a stream of data releases, thus acting as a central point for  the collection, curation, and distribution of the data.  The proliferation of cloud-based tools and open-source software have now reduced these roadblocks considerably.  We argue that by linking together the tools described above, researchers can now produce, curate, and distribute Versioned Data much more easily and cheaply than in the past.

@@

Because \texttt{git} was built for tracking changes in code, there are aspects of how it behaves that are not natural to apply to  data \citep{Perkel-2016}.  For example, because it tracks changes line-by-line rather than cell-by-cell, there are inefficiencies in how the software stores the changes.
There is rapid current work on new implementations which will be faster computationally, smaller in file size, flexible with respect to data structures, and thus allow for larger datasets to be versioned efficiently \citep{Fli, Dat}.
% Add CKAN, rdrop2 and Git-LFS, conceptually at least, and say something useful about the future
